A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-05 tex/codatadp.tex
more

Blah.

  * briefly: datatypes and why they are useful
    * Hoare's recursive data structures
    * comparatively easy to reason about and automate
    * common in practice, esp. in computer science applications

    (terminology: freely-generated, inductive, algebraic, ..., sometimes
    with different meanings; we'll clarify below what we use)

@@

@@

A (co)datatype specification consists of $l$~mutually recursive types which are
either all datatypes or all codatatypes. Each datatype $\delta$ is equipped by
$m \ge 1$ constructors, and each constructor for $\delta$ takes zero or more
arguments and returns a $\delta$ value. The argument types must be either
plain, be or be one of the newly introduced (co)datatypes. To every
argument corresponds a selector. The introduced names for the (co)datatypes, the
constructors, and the selectors must be distinct and different from already
introduced ones.%
\footnote{It can be convenient to specify the same selector
for several constructors associated with the same (co)data\-type,
as long as the argument types coincide. % \cite[Section~3]{blanchette-et-al-2014-codata}.
However, this is disallowed by SMT-LIB 2, so we do not consider it here.}
Schematically, using a syntax similar to Standard ML, Haskell,
or SMT-LIB:
%
\[
\begin{aligned}[t]
(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}=\, \const{C}_{1}\rlap{${}_{1}$}\phantom{{}_{m}}(\bigl[\const{s}_{11}^1{:}\bigr]\; \tau_{11}^1, \,\ldots,\, \bigl[\const{s}_{11}^{n_{11}}{:}\bigr]\; \tau_{11}^{n_{11}}) & \\
           & {}\;\phantom{=\,}\;{} \smash{\,\,\,\vdots} \\[-\jot] % {}\;\phantom{=}\llap{\ensuremath{\mid}}\; \cdots \\
           & {}\;\phantom{=\,}\llap{\ensuremath{\,\mid\,}}\; \const{C}_{1m}(\bigl[\const{s}_{1m}^1{:}\bigr]\; \tau_{1m}^1, \,\ldots,\, \big[\const{s}_{1m}^{n_{1m}}{:}\big]\; \tau_{1m}^{n_{1m}}) \\[-1\jot]
   \smash{\vdots\,\,\,} \\[-1\jot]
  \keyw{and}\; \,\delta_l & =\, \ldots
\end{aligned}
\]
%
with
$\const{C}_{i\negvthinspace j} : \tau_{i\negvthinspace j}^1\times\cdots\times\tau_{i\negvthinspace j}^{\,k_{\smash{i\negvthinspace j}}} \to \delta_i$
and $\const{s}_{i\negvthinspace j}^{\,k} : \delta_i \to \tau_{i\negvthinspace j}^{\,k}$. Defaults are assumed for
the selector names if they are omitted.

>>>>

Freely generated datatypes are widely used both in functional programs and in
logical specifications. They can be implemented efficiently and enjoy nice
properties that can be exploited in automated reasoners. Although they have
their uses in mathematics, they are especially prominent in computer science
applications.

However, because datatype values correspond to ground terms, they can only
represent finite objects. For example, the datatype of natural numbers
constructed by $\const{Z} : \ty{nat}$ and $\const{S} : \ty{nat} \to \ty{nat}$
only allow values of the form $\const{S}(\ldots(\const{S}(\const{Z}))\ldots)$.
To represent infinite values such as $\const{S}(\const{S}(\ldots))$, the natural
choice is to turn to codatatypes, the coinductive dual of datatypes.

  * are sometimes perceived as esoteric
  * but they have their uses where they are nearly irreplacable
  * for example, CompCert and JinjaThreads

  * codatatype approach unifies the finite and infinite case; it is often more
    convenient to have
      codatatype enat = EZ | ES enat
    than
      datatype enat = EZ | ES enat | EI
    with $\infty$ represented by $ES(ES(\ldots))$. Notice that adding
    $\infty$, as expected

  * also can be understood very easily:
    intuitively, they consist of all ground constructor terms, finite and infinite
    and only those
  * one implication is that if $m$ equals $ES(n)$ and $n$ equals $ES(m)$, necessarily
    $m$ and $n$ must be equal.

  * perhaps the most commonly used codatatype is that of lazy lists or sequences.
    using a syntax similar to Standard ML, Haskell, or SMT-LIB

$\mu$ notation

@@

%    (terminology: freely-generated, inductive, algebraic, ..., sometimes
%    with different meanings; we'll clarify below what we use)

@@

A (co)datatype specification consists of $l$~mutually recursive types which are
either all datatypes or all codatatypes. Each datatype $\delta$ is equipped by
$m \ge 1$ constructors, and each constructor for $\delta$ takes zero or more
arguments and returns a $\delta$ value. The argument types must be either
plain, be or be one of the newly introduced (co)datatypes. To every
argument corresponds a selector. The introduced names for the (co)datatypes, the
constructors, and the selectors must be distinct and different from already
introduced ones.%
\footnote{It can be convenient to specify the same selector
for several constructors associated with the same (co)data\-type,
as long as the argument types coincide. % \cite[Section~3]{blanchette-et-al-2014-codata}.
However, this is disallowed by SMT-LIB 2, so we do not consider it here.}
Schematically:
%
\[
\begin{aligned}[t]
(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}=\, \const{C}_{1}\rlap{${}_{1}$}\phantom{{}_{m}}(\bigl[\const{s}_{11}^1{:}\bigr]\; \tau_{11}^1, \,\ldots,\, \bigl[\const{s}_{11}^{n_{11}}{:}\bigr]\; \tau_{11}^{n_{11}}) & \\
           & {}\;\phantom{=\,}\;{} \smash{\,\,\,\vdots} \\[-\jot] % {}\;\phantom{=}\llap{\ensuremath{\mid}}\; \cdots \\
           & {}\;\phantom{=\,}\llap{\ensuremath{\,\mid\,}}\; \const{C}_{1m}(\bigl[\const{s}_{1m}^1{:}\bigr]\; \tau_{1m}^1, \,\ldots,\, \big[\const{s}_{1m}^{n_{1m}}{:}\big]\; \tau_{1m}^{n_{1m}}) \\[-1\jot]
   \smash{\vdots\,\,\,} \\[-1\jot]
  \keyw{and}\; \,\delta_l & =\, \ldots
\end{aligned}
\]
%
with
$\const{C}_{i\negvthinspace j} : \tau_{i\negvthinspace j}^1\times\cdots\times\tau_{i\negvthinspace j}^{\,k_{\smash{i\negvthinspace j}}} \to \delta_i$
and $\const{s}_{i\negvthinspace j}^{\,k} : \delta_i \to \tau_{i\negvthinspace j}^{\,k}$. Defaults are assumed for
the selector names if they are omitted.
